---
title: "ROI - R Optimization Infrastructure"
output: html_document
---

<style type="text/css">
.title {
  display: none;
}

#getting-started img {
  margin-right: 10px;
}

</style>

<div class="row" style="padding-top: 30px;">

# **ROI** 

The R Optimization Infrastructure (ROI) package provides an extensible 
infrastructure to model linear, quadratic, conic and general nonlinear 
optimization problems in a consistent way. 

```{r, eval = FALSE}
library(ROI)
```

```{r, echo = FALSE}
Sys.setenv(ROI_LOAD_PLUGINS = "FALSE")
suppressPackageStartupMessages(library(ROI))
```

Furthermore, the infrastructure administers many different solvers, 
reformulations, problem collections and functions to read and write 
optimization problems in various formats.

# Extensions

**ROI** provides the modeling capabilities and manages the plugins, the plugins
add the solvers to **ROI**.


### Plugins

````{r, results = 'asis', message = FALSE}
plugins <- ROI_available_solvers()[,c("Package", "Repository")]
plugins <- aggregate(Repository ~ Package,  data = plugins, 
                     FUN = paste, collapse = ", ")
knitr::kable(plugins, row.names = TRUE)
```

<!-- 
| Plugin                        | Objective | Constraints | Bounds | Types | Repository |
| ----------------------------- | --------- | ----------- | ------ | ----- | ---------- |
| ROI.plugin.alabama            | F         |             |        |       |            |
| ROI.plugin.clp                | L         |             |        |       |            |
| ROI.plugin.cplex              | Q         |             |        |       |            |
| ROI.plugin.deoptim            | F         |             |        |       |            |
| ROI.plugin.ecos               | L         |             |        |       |            |
| ROI.plugin.glpk               | L         |             |        |       |            |
| ROI.plugin.ipop               | Q         |             |        |       |            |
| ROI.plugin.lpsolve            | L         |             |        |       |            |
| ROI.plugin.msbinlp            | L         |             |        |       |            |
| ROI.plugin.neos               | X, L, Q   | X, L, Q     |        |       |            |
| ROI.plugin.nloptr             |           |             |        |       |            |
| ROI.plugin.optimx             |           |             |        |       |            |
| ROI.plugin.qpoases            |           |             |        |       |            |
| ROI.plugin.quadprog           |           |             |        |       |            |
| ROI.plugin.scs                |           |             |        |       |            |
| ROI.plugin.symphony           |           |             |        |       |            |
-->

### Models

- [ROI.models.globalOptTests](https://CRAN.R-project.org/package=ROI.models.globalOptTests)
- [ROI.models.miplib](https://CRAN.R-project.org/package=ROI.models.miplib)
- [ROI.models.netlib](https://CRAN.R-project.org/package=ROI.models.netlib)



# Installation
The [Installation](installation.html) page contains information to assist with the installation 
of **ROI** and its companion packages.

# Additional resources
Additional to this web page there exist several resources:

- [ROI - Manual](https://CRAN.R-project.org/package=ROI/ROI.pdf)
- [ROI - Working paper](http://epub.wu.ac.at/5858/)
- [ROI.plugin.neos - Introduction](https://CRAN.R-project.org/package=ROI.plugin.neos/vignettes/ROI.plugin.neos_Introduction.pdf)

# Planed extensions
### Planed plugins (**ROI.plugin.***)   

  - [OSQP](https://osqp.org/) there exist already the **R** package 
    [osqp](https://cran.r-project.org/package=osqp),
    [rosqp](https://cran.r-project.org/package=rosqp)
  - [Couenne](https://projects.coin-or.org/Couenne)

### Planed models (**ROI.model.***)   

  - [QPLIB](http://qplib.zib.de/)

# Contribute
There are several possible ways to contribute to the **ROI** project.

- Since **ROI** was designed to allow decentralized development anyone is 
  free and invited to extend **ROI** by creating a new plugin or model collection.
- The source code of this page is stored on [GitHub](https://github.com/FlorianSchwendinger/ROI_docs) modifications
  to this page can be suggested by raising an issue or creating a pull request.
- If you find **ROI** useful we would be happy to add your particular use
  case to the [Use Cases](use_cases.html) page. To do so you can again raise
  and issue or create a pull request on [GitHub](https://github.com/FlorianSchwendinger/ROI_docs), or send an e-mail 
  to [roi[at]shared-files.de](mailto:roi@shared-files.de). 
  The use case should be either written in **Sweave**, **knitr** or **R Markdown**
  and contain an author and a license (e.g. [GNU General Public License version 3](https://en.wikipedia.org/wiki/GNU_General_Public_License)).

# Citation
```{asis}
Theußl, Stefan and Schwendinger, Florian and Hornik, Kurt (2017) ROI: The R Optimization Infrastructure Package. Research Report Series / Department of Statistics and Mathematics, 133. WU Vienna University of Economics and Business, Vienna.
```
```
@techreport{epubwu5858,
  number = {133},
  month = {October},
  author = {Stefan Theu{\ss}l and Florian Schwendinger and Kurt Hornik},
  address = {Vienna},
  title = {ROI: The R Optimization Infrastructure Package},
  type = {Research Report Series / Department of Statistics and Mathematics},
  institution = {WU Vienna University of Economics and Business},
  year = {2017},
  keywords = {optimization, mathematical programming,
    linear programming, quadratic programming, convex
    programming, nonlinear programming, mixed integer programming, R},
  url = {http://epub.wu.ac.at/5858/},
  abstract = {Optimization plays an important role in many methods routinely 
    used in statistics, machine learning and data science. Often, 
    implementations of these methods rely on highly specialized (non-reusable) 
    optimization algorithms. However, in many instances recent advances, in 
    particular in the field of convex optimization, make it possible to 
    conveniently and straightforwardly use modern solvers (reusable) instead
    with the advantage of enabling broader usage scenarios and thus
    promoting reusability. This paper introduces the ROI
    (R Optimization Infrastructure) package which provides an
    extensible infrastructure to model linear, quadratic, conic and general
    nonlinear optimization problems in a consistent way. Furthermore, the
    infrastructure administers many different solvers, reformulations,
    problem collections and functions to read and write optimization
    problems in various formats.}
```

</div>

