---
title: "Installation"
---

The following contains additional information for the installation 
of **ROI** and its companion packages.

# **ROI**
Since **ROI** has no dependencies to external system and only imports 
libraries written in base **R**, the installation via
`install.packages("ROI")` should be sufficient on any platform. 

# **ROI plug-ins**

## **ROI.plugin.alabama**
```{r, eval = FALSE}
install.packages("ROI.plugin.alabama")
```

## **ROI.plugin.cbc**

### Unix

[**ROI.plugin.cbc**](https://github.com/dirkschumacher/ROI.plugin.cbc)
it self doesn't contain compiled code but it imports **rcbc**.
The [**rcbc**](https://github.com/dirkschumacher/rcbc) package requires 
the [cbc](https://projects.coin-or.org/Cbc) and 
the [clp](https://projects.coin-or.org/Clp) library to be installed before
the installation of **rcbc**. Depending on the distribution, 
these libraries may be installed via one of the following commands.


#### Debian, Ubuntu, Mint and other Debian based distributions

```{bash, eval = FALSE}
sudo apt-get install coinor-libcbc-dev coinor-libclp-dev
```

#### Red Hat Enterprise Linux, Fedora, CentOS and similar distributions

```{bash, eval = FALSE}
sudo yum install coin-or-Cbc-devel coin-or-Clp-devel
```

#### macOS
```{bash, eval = FALSE}
brew tap coin-or-tools/coinor
brew install cbc
```

After the installation of the required libraries
**rcbc** and **ROI.plugin.cbc** may be installed as usual.

```{r, eval = FALSE}
remotes::install_github("dirkschumacher/rcbc")
remotes::install_github("dirkschumacher/ROI.plugin.cbc")
```

### Windows
```{r, eval = FALSE}
devtools::install_github("dirkschumacher/rcbc")
devtools::install_github("dirkschumacher/ROI.plugin.cbc")
```

## **ROI.plugin.clp**
```{r, eval = FALSE}
install.packages("ROI.plugin.clp")
```

## **ROI.plugin.cplex**
```{r, eval = FALSE}
install.packages("ROI.plugin.cplex")
```

## **ROI.plugin.deoptim**
```{r, eval = FALSE}
install.packages("ROI.plugin.deoptim")
```

## **ROI.plugin.ecos**
```{r, eval = FALSE}
install.packages("ROI.plugin.ecos")
```

## **ROI.plugin.glpk**
[**ROI.plugin.glpk**](https://cran.r-project.org/package=ROI.plugin.glpk)
it self doesn't contain compiled code but it imports **Rglpk**.
The [**Rglpk**](https://cran.r-project.org/package=Rglpk) package requires
the linear programming kit - development files to be installed before
the installation of **Rglpk**. Depending on the distribution, 
the linear programming kit - development files may be installed via one
of the following commands.

### Unix

#### Debian, Ubuntu, Mint and other Debian based distributions

```{bash, eval = FALSE}
sudo apt-get install libglpk-dev
```

#### Red Hat Enterprise Linux, Fedora, CentOS and similar distributions

```{bash, eval = FALSE}
yum install glpk-devel
```

#### macOS
```{bash, eval = FALSE}
brew install glpk
```

After the installation of the linear programming kit - development files
**Rglpk** and **ROI.plugin.glpk** may be installed as usual.

```{r, eval = FALSE}
install.packages("ROI.plugin.glpk")
```

### Windows
```{r, eval = FALSE}
install.packages("ROI.plugin.glpk")
```

## **ROI.plugin.gurobi**
**ROI.plugin.gurobi** imports the **gurobi** package which is available at
[http://www.gurobi.com](http://www.gurobi.com/). Since the **gurobi** package
is not on `CRAN` **ROI.plugin.gurobi** is also not available on `CRAN`.
```{r, eval = FALSE}
remotes:::install_github("fl0sch/ROI.plugin.gurobi")
```

## **ROI.plugin.ipop**
```{r, eval = FALSE}
install.packages("ROI.plugin.ipop")
```

## **ROI.plugin.lpsolve**
```{r, eval = FALSE}
install.packages("ROI.plugin.lpsolve")
```

## **ROI.plugin.mosek**
**ROI.plugin.mosek** imports the **Rmosek** package which is available at
[http://www.mosek.com](http://www.mosek.com/). There is also an 
**Rmosek** version on `CRAN` but since the **Rmosek** version on `CRAN`
is very old we recommend to install the version available from
[http://www.mosek.com](http://www.mosek.com/).

```{r, eval = FALSE}
remotes:::install_github("fl0sch/ROI.plugin.mosek")
```

## **ROI.plugin.msbinlp**
```{r, eval = FALSE}
install.packages("ROI.plugin.msbinlp")
```

## **ROI.plugin.neos**

### Unix

[**ROI.plugin.neos**](https://cran.r-project.org/package=ROI.plugin.neos)
it self doesn't contain compiled code but it imports xml2 and curl.
The [**xml2**](https://cran.r-project.org/package=xml2) package requires that
the `libxml2` development library is installed. Depending on the distribution, 
this library may is installed via one of the following commands.

### Unix

#### Debian, Ubuntu, Mint and other Debian based distributions
```{bash, eval = FALSE}
sudo apt-get install libxml2-dev 
```

#### Red Hat Enterprise Linux, Fedora, CentOS and similar distributions
```{r, eval = FALSE}
sudo yum install libxml2-devel
```

#### macOS

After the installation of the the `libxml2` development libraries
**xml2** may be installed as usual.

```{r, eval = FALSE}
install.packages("xml2")
```

The [**curl**](https://cran.r-project.org/package=curl) package requires that
the `curl` development library is installed. Depending on the distribution, 
this library may is installed via one of the following commands.

### Unix

#### Debian, Ubuntu, Mint and other Debian based distributions
```{bash, eval = FALSE}
sudo apt-get install libcurl4-openssl-dev
```

#### Red Hat Enterprise Linux, Fedora, CentOS and similar distributions
```{r, eval = FALSE}
sudo yum install libcurl-devel
```

#### macOS

After the installation of the the `libxml2` development libraries
**xmle** can be installed.
```{r, eval = FALSE}
install.packages("curl")
```

After the **curl** and **xml2** packages are installed **ROI.plugin.neos**
can be installed.
```{r, eval = FALSE}
install.packages("ROI.plugin.neos")
```

### Windows

```{r, eval = FALSE}
install.packages("ROI.plugin.neos")
```

## **ROI.plugin.nloptr**
```{r, eval = FALSE}
install.packages("ROI.plugin.nloptr")
```

## **ROI.plugin.optimx**
```{r, eval = FALSE}
install.packages("ROI.plugin.optimx")
```

## **ROI.plugin.qpoases**
```{r, eval = FALSE}
install.packages("ROI.plugin.qpoases")
```

## **ROI.plugin.quadprog**
```{r, eval = FALSE}
install.packages("ROI.plugin.quadprog")
```

## **ROI.plugin.scs**

```{r, eval = FALSE}
install.packages("ROI.plugin.scs")
```

## **ROI.plugin.symphony**

### Unix

[**ROI.plugin.symphony**](https://cran.r-project.org/package=ROI.plugin.symphony)
it self doesn't contain compiled code but it imports **Rsymphony**.
The [**Rsymphony**](https://cran.r-project.org/package=Rsymphony) package requires
the SYMPHONY libraries and headers to be installed before
the installation of **Rsymphony**. Depending on the distribution, 
the SYMPHONY libraries and headers may be installed via one
of the following commands.

#### Debian, Ubuntu, Mint and other Debian based distributions

```{bash, eval = FALSE}
sudo apt install coinor-libsymphony-dev
```

#### Red Hat Enterprise Linux, Fedora, CentOS and similar distributions

```{bash, eval = FALSE}
sudo yum install coin-or-SYMPHONY-devel
```

#### macOS

```{bash, eval = FALSE}
brew tap coin-or-tools/coinor
brew install symphony
```

After the installation of the SYMPHONY libraries and headers
**ROI.plugin.symphony** may be installed as usual.

```{r, eval = FALSE}
install.packages("ROI.plugin.symphony")
```

### Windows

```{r, eval = FALSE}
install.packages("ROI.plugin.symphony")
```

# **ROI models**

## **ROI.models.netlib**
```{r, eval = FALSE}
install.packages("ROI.models.netlib")
```

## **ROI.models.miplib**
```{r, eval = FALSE}
install.packages("ROI.models.miplib")
library(ROI.models.miplib)
miplib_download_metainfo("http://miplib2010.zib.de/download/miplib2010_all.solu")
miplib_download_benchmark("http://miplib2010.zib.de/download/miplib2010-1.1.3-benchmark.tgz",
    quiet = FALSE)
```



