---
title: "Modeling with **ROI**"
---

```{r, echo = FALSE}
is_gurobi_installed <- any(installed.packages()[, "Package"] == "ROI.plugin.gurobi")
```

```{r}
Sys.setenv(ROI_LOAD_PLUGINS = "FALSE")
library(ROI)
library(ROI.plugin.glpk)
library(ROI.plugin.qpoases)
library(ROI.plugin.ecos)
library(ROI.plugin.scs)
library(ROI.plugin.alabama)
library(ROI.plugin.lpsolve)
```

```{r, eval = is_gurobi_installed}
library(ROI.plugin.gurobi)
```

# Constructors

## Optimization Problem 
```{r}
str(OP)
```

## Objective
### Linear objective
```{r}
str(L_objective)
```

$$1 x + 2 y + 3 z$$

```{r}
lo <- L_objective(c(1, 2, 3), c("x", "y", "z"))
```

### Quadratic objective

```{r}
str(Q_objective)
```

$$\frac{1}{2} (x_1^2 + x_2^2 + x_3^2) + 1 x_1 + 2x_2 + 3x_3$$

```{r}
qo <- Q_objective(diag(3), c(1, 2, 3), c("x_1", "x_2", "x_3"))
```

### Functional objective}
```{r}
str(F_objective)
```

$$x_1^2 + x_2^2$$

```{r}
fo <- F_objective(F = function(x) sum(x^2), n = 2, 
                  G = function(x) 2*x, names = c("x_1", "x_2"))
```

## Constraints
### Linear constraints
```{r}
str(L_constraint)
```

\[
\nonumber
\begin{array}{rrrrrrr}
3 x & + & 4 y & + & 1 z & \leq & 90 \\
1 x & + & 0 y & + & 2 z & \geq & 5  \\
1 x & + & 1 y & + & 0 z &   =  & 2
\end{array}
\]

```{r}
lc <- L_constraint(L = rbind(c(3, 4, 1), c(1, 0, 2), c(1, 1, 0)),
                   dir = c("<=", ">=", "=="), rhs = c(90, 5, 2),
                   names = c("x", "y", "z"))
```

### Quadratic constraints
```{r}
str(Q_constraint)
```

$$\frac{1}{2} (x^2 + y^2) + 1x + 2y \leq 3$$

```{r}
qc1 <- Q_constraint(Q = diag(2), L = 1:2, dir = "<=", 
                    rhs = 3, names = c("x", "y"))
```

\[
\begin{array}
x^2 + y^2 + 3x + 1y & \leq & 3 \nonumber \\
x + y & \leq & 4 \nonumber \\
\frac{1}{2} (3x^2 + 3y^2 + 2xy) + 2x + 5y & \leq & 9 \nonumber
\end{array}
\]

```{r}
qc2 <- Q_constraint(Q = list(diag(2, 2), NULL, matrix(c(3, 1, 1, 3), 2)), 
                    L = rbind(c(3, 1), c(1, 1), c(2, 5)),
                    dir = c("<=", "<=", "<="),
                    rhs = c(3, 4, 9), names = c("x", "y"))
```

### Functional constraints
```{r}
str(F_constraint)
```

\[
\begin{array}
x^2 & \leq & 2 \nonumber \\
y^2 & \leq & 4 \nonumber
\end{array}
\]

```{r}
fc1 <- F_constraint(F = function(x) x^2, dir = c("<=", "<="), rhs = c(2, 4),
                    J = function(x) diag(x = 2, nrow = 2) * x,
                    names = c("x", "y"))
```

or equivalently

```{r}
fc2 <- F_constraint(F = list(function(x) x[1]^2, function(x) x[2]^2), 
                    dir = c("<=", "<="), rhs = c(2, 4),
                    J = list(function(x) rbind(c(2, 0) * x), 
                             function(x) rbind(c(0, 2) * x)),
                    names = c("x", "y"))

## TODO: create an example
## x <- OP(L_objective(c(1, 1)), fc1, maximum=TRUE)
## x <- OP(L_objective(c(1, 1)), fc2, maximum=TRUE)
## solution(ROI_solve(x, start=c(0, 1), solver="alabama"))
```

## Bounds
### Variable bounds
By default the variable bounds are set to  $0 \leq x_i \leq \infty \text{ for all } i = 1, ..., n.$

```{r}
str(V_bound)
```
$$ -3 \leq x_1 \leq 3, \ -\infty \leq x_2 \leq 7, \ -9 \leq x_3 \leq \infty $$

```{r}
vb <- V_bound(li = 1:3, ui = 1:3, lb = c(-3, -Inf, -9), ub = c(3, 7, Inf))
```

# Basic examples
## Linear programming (LP)
### Example 1
\[
\begin{array}{rrrrr}
\text{minimize}
& 7 x_1 & + & 8 x_2 \\
\text{subject to}
& 3 x_1 & + & 4 x_2 &  =   9 \\
& 2 x_1 & + & 1 x_2 & \geq 3
\end{array}
\]

$$ -100 \leq x_1, x_2, \leq 100$$

```{r}
lp  <- OP(objective = L_objective(c(7, 8), names=c("x", "y")),
          constraints = L_constraint(L = rbind(c(3, 4), c(2, 1)), 
                                     dir = c("==", ">="), rhs = c(9, 3)),
          bounds = V_bound(li = 1:2, ui = 1:2, 
                           lb = c(-100, -100), ub = c(100, 100)))
ROI_applicable_solvers(lp)
(sol <- ROI_solve(lp, solver = "glpk"))
solution(sol)
```

The solution can be accessed via the function \code{solution},
where    

```{r}
solution(sol)
solution(sol, type = "primal")
```

gives the primal solution, 

```{r}
solution(sol, type = "dual")
```

the dual solution, 

```{r}
solution(sol, type = "msg")
```

the original message returned from the solver.


### **Example 2**
$$\begin{array}{rrrrrrr}
\text{maximize}
& 7 x_1 & + & 3 x_2 & + & 1 x_3 & \\
\text{subject to}
& 6 x_1 & + & 4 x_2 & + & 5 x_3 & \leq 60 \\
& 8 x_1 & + &   x_2 & + & 2 x_3 & \leq 80 \\
& 9 x_1 & + & 1 x_2 & + & 7 x_3 & \leq 70 
\end{array}
$$
$$x_1, x_2, x_3 \geq 0$$

```{r}
lp <- OP(objective = L_objective(c(7, 1, 3), c("x", "y", "z")),
         constraints = L_constraint(L = rbind(c(6, 4, 5), c(8, 0, 2), c(9, 1, 7)),
                                    dir = c("<=", "<=", "<="),
                                    rhs = c(60, 80, 70)),
         maximum = TRUE)
(sol <- ROI_solve(lp))
solution(sol)
```

## Mixed integer linear programming (MILP)
$$\begin{array}{rrrrrrr}
\text{maximize}
& 7 x_1 & + & 3 x_2 & + & 1 x_3 & \\
\text{subject to}
& 6 x_1 & + & 4 x_2 & + & 5 x_3 & \leq 60 \\
& 8 x_1 & + &   x_2 & + & 2 x_3 & \leq 80 \\
& 9 x_1 & + & 1 x_2 & + & 7 x_3 & \leq 70 
\end{array}
$$
$$x_1, x_3 \in \mathbb{Z}_{\geq 0}$$
$$x_2 \geq 0$$

```{r}
A <- rbind(c(6, 4, 5), c(8, 0, 2), c(9, 1, 7))
milp <- OP(objective = L_objective(c(7, 1, 3), c("x", "y", "z")),
           constraints = L_constraint(L = rbind(c(6, 4, 5), c(8, 0, 2), c(9, 1, 7)),
                                      dir = c("<=", "<=", "<="),
                                      rhs = c(60, 80, 70)),
           types = c("I", "C", "I"), 
           maximum = TRUE)
(sol <- ROI_solve(milp))
solution(sol)
```

## Quadratic programming
### Quadratic objective with linear constraints (QP)

$$
\begin{array}
\text{minimize}   & & x_1 & + & 2x_2 & + & 3x_3 &  +   & \frac{1}{2} (x_1^2 + x_2^2 + x_3^2)  \\
\text{subject to} & & x_1 & + &  x_2 &   &      & \geq & 1 \nonumber \\
                  & &     &   &  x_2 & + &  x_3 & \geq & 2 \nonumber \\
                  & & x_1 &   &      & + &  x_3 & \geq & 3 \nonumber \\
\end{array}
$$

```{r}
qp <- OP(Q_objective(diag(3), c(1, 2, 3), c("x", "y", "z")),
         L_constraint(L = rbind(c(1, 1, 0), c(0, 1, 1), c(1, 0, 1)), 
                      dir = c(">=", ">=", ">="), rhs = c(1, 2, 3)))
(sol <- ROI_solve(qp, solver = "qpoases"))
solution(sol)
```

### Quadratic objective with quadratic constraints (QCQP)
$$
\text{maximize} \ \ 90 x_1 + 110 x_2 + 160 x_3 - \frac{1}{2} (x_1^2 + x_2^2 + x_3^2)
$$
$$
\begin{array}{rrrrr}
\text{subject to} & x_1^2 + x_2^2 + 4 x_3     & \leq & 4 & \\
                  & x_2^2 + x_3^2 + x_1 + x_3 & \leq & 3 & \\
                  & x_1^2 + x_3^2 + 2 x_1 x_3 & \leq & 2 & \\
                  & x_1, x_2, x_3 \geq 0      &      &   &
\end{array}
$$

```{r}
qcqp <- OP(Q_objective(-diag(3), c(90, 110, 160), c("x", "y", "z")),
           Q_constraint(Q = list(rbind(c(2, 0, 0), c(0, 2, 0), c(0, 0, 0)),
                                 rbind(c(0, 0, 0), c(0, 2, 0), c(0, 0, 2)),
                                 rbind(c(2, 0, 2), c(0, 0, 0), c(2, 0, 2))),
                        L = rbind(c(0, 0, 4), c(1, 0, 1), c(0, 0, 0)),
                        dir = rep("<=", 3), rhs = c(4, 3, 2)),
           maximum = TRUE)
```
```{r, eval = is_gurobi_installed}
(sol <- ROI_solve(qcqp, solver = "gurobi"))
solution(sol)
```

or equivalently

```{r, eval = is_gurobi_installed}
qc1 <- Q_constraint(Q = rbind(c(2, 0, 0), c(0, 2, 0), c(0, 0, 0)),
                    L = c(0, 0, 4), dir = "<=", rhs = 4)
qc2 <- Q_constraint(Q = rbind(c(0, 0, 0), c(0, 2, 0), c(0, 0, 2)),
                    L = c(1, 0, 1), dir = "<=", rhs = 3)
qc3 <- Q_constraint(Q = rbind(c(2, 0, 2), c(0, 0, 0), c(2, 0, 2)),
                    L = NULL, dir = "<=", rhs = 2)
qcqp <- OP(Q_objective(-diag(3), c(90, 110, 160), c("x", "y", "z")),
           c(qc1, qc2, qc3), maximum = TRUE)

(sol <- ROI_solve(qcqp, solver = "gurobi"))
solution(sol)
```

Check how tight the bounds are.

```{r}
sapply(as.function(constraints(qcqp)), function(F) F(solution(sol)))
```

## Conic programming (CP)
### Second-order cone programming (SOCP)
$$\text{maximize } \ \ x + y$$
$$\text{subject to } \ \ \sqrt{x^2 + y^2} \leq \sqrt{2}$$
$$x, y \geq 0$$

```{r}
socp1 <- OP(objective = L_objective(c(1, 1), names = c("x", "y")),
            constraints = C_constraint(rbind(c(0, 0), c(-1, 0), c(0, -1)),
                                       cones = K_soc(3), 
                                       rhs = c(sqrt(2), 0, 0)),
            maximum = TRUE)
(sol <- ROI_solve(socp1))
solution(sol)
```

### Positive semidefinite programming (SDP)
$$
\begin{array}{rl}
\text{minimize}   & x_1 + x_2 - x_3 \\
\text{subject to} &
x_1
\begin{pmatrix}
10 & 3 \\
3  & 10
\end{pmatrix}
+
x_2
\begin{pmatrix}
6  & -4 \\
-4 & 10
\end{pmatrix}
+
x_3
\begin{pmatrix}
8  &  1 \\
1  &  6
\end{pmatrix}
\preceq
\begin{pmatrix}
 16  & -13 \\
-13  &  60
\end{pmatrix}
\\
& x_1, x_2, x_3 \geq 0
\nonumber
\end{array}
$$

```{r}
A1 <- rbind(c(10, 3), c(3, 10))
A2 <- rbind(c(6, -4), c(-4, 10))
A3 <- rbind(c(8, 1), c(1, 6))
A4 <- rbind(c(16, -13), c(-13, 60))
psd <- OP(objective = L_objective(c(1, 1, -1)),
          constraints = C_constraint(L = vech(A1, A2, A3), 
                                     cones = K_psd(3),
                                     rhs = vech(A4)))
(sol <- ROI_solve(psd))
solution(sol)
as.matrix(solution(sol, "psd")[[1]])
```

The following example taken from the [CVXOPT homepage](http://cvxopt.org/userguide/coneprog.html#semidefinite-programming).

$$
\begin{array}{rl}
\text{minimize}   & x_1 - x_2 + x_3 \\
\text{subject to} &
x_1 \
\begin{pmatrix}
 -7 & -11 \\
-11 &   3
\end{pmatrix}
+ x_2 \
\begin{pmatrix}
  7 & -18 \\
-18 &   8
\end{pmatrix}
+ x_3 \
\begin{pmatrix}
 -2 &  -8 \\
 -8 &   1
\end{pmatrix}
\ \preceq \
\begin{pmatrix}
 33 &  -9 \\
 -9 &  26
\end{pmatrix} 
\ \\
&
x_1 \
\begin{pmatrix}
-21 & -11 &   0 \\
-11 &  10 &   8 \\
  0 &   8 &   5
\end{pmatrix}
+ x_2 \
\begin{pmatrix}
  0 &  10 &  16 \\
 10 & -10 & -10 \\
 16 & -10 &   3
\end{pmatrix}
+ x_3 \
\begin{pmatrix}
 -5 &   2 & -17 \\
  2 &  -6 &   8 \\
-17 &   8 &   6
\end{pmatrix}
\ \preceq \
\begin{pmatrix}
 14 &   9 &  40 \\
  9 &  91 &  10 \\
 40 &  10 &  15
\end{pmatrix}
\ \\
& x_1, x_2, x_3 \in \mathbb{R}
\end{array}
\nonumber
$$

```{r}
obj <- c(1, -1, 1)
A1 <- matrix(c(-7, -11, -11,  3), 2)
A2 <- matrix(c( 7, -18, -18,  8), 2)
A3 <- matrix(c(-2,  -8,  -8,  1), 2)
A4 <- matrix(c(33,  -9,  -9, 26), 2)
B1 <- matrix(c(-21, -11,  0, -11,  10,   8,  0,    8, 5), 3)
B2 <- matrix(c(  0,  10,  16, 10, -10, -10,  16, -10, 3), 3)
B3 <- matrix(c( -5,   2, -17,  2,  -6,   8, -17,   8, 6), 3)
B4 <- matrix(c( 14,   9,  40,  9,  91,  10,  40,  10,15), 3)
rhs <- c(vech(A4), vech(B4))

psd <- OP(objective = obj,
          constraints = C_constraint(L = rbind(vech(A1, A2, A3), vech(B1, B2, B3)), 
                                     cones = K_psd(c(3, 6)), rhs = rhs),
          bounds = V_bound(li=1:3, lb=rep(-Inf, 3)))

(sol <- ROI_solve(psd, solver = "scs"))
solution(sol)
lapply(solution(sol, type="psd"), as.matrix)
```

### Exponential cone
$$
\begin{array}
\mathcal{K}_{exp} &=&
            \{(x, y, z) | y  > 0, y e^{\frac{x}{y}} \leq z \} \cup
            \{(x, 0, z) | x \leq 0, z \geq 0\} \\
\mathcal{K}_{exp}^* &=&
            \{(u, v, w) | u < 0, -ue^\frac{v}{u} \leq ew\} \cup
            \{(0, v, w) | v \geq 0, w \geq 0\}
\end{array}
$$

#### Primal exponential cone
$$
\begin{array}
\text{maximize}   & x + y + z \\
\text{subject to} & y   e^{\frac{x}{y}} \leq z \\
& x \geq 0, y > 0, z \in [0, e]
\end{array}
$$

```{r}
expp <- OP(objective = L_objective(c(1, 1, 1)),
           constraints = C_constraint(diag(-1, 3), cones = K_expp(1), 
                                      rhs = rep(0, 3)),
           bounds = V_bound(li = 2, lb = 1e-12, ui = 3, ub = exp(1)),
           maximum = TRUE)
(sol <- ROI_solve(expp, tol=1e-8, solver="scs"))
solution(sol)
```

#### Dual exponential cone
$$
\begin{array}
\text{minimize}   & u + v + w \\
\text{subject to} & -u e^{\frac{v}{u}} \leq ew \\
& u \in [-1, 0], y, z \geq 0
\end{array}
$$

```{r}
expd <- OP(objective = L_objective(c(1, 1, 1), names = c("u", "v", "w")),
           constraints = C_constraint(diag(x=-1, 3), cones = K_expd(1), 
                                      rhs = rep(0, 3)), 
           bounds = V_bound(li = 1, lb = -1, ui = 1, ub = 0, nobj = 3L))
(sol <- ROI_solve(expd))
solution(sol)
```


### Power cone
$$
\begin{array}
\mathcal{K}_{pwr}^\alpha &=& 
                \{(x, y, z) | x^\alpha y^{1-\alpha} \geq |z|, x, y \geq 0\},
                \text{ where } \alpha \in [0, 1] \\
\left(\mathcal{K}_{pwr}^\alpha\right)^* &=&
                \left\{(u, v, w) | \left(\frac{u}{a}\right)^\alpha \left(\frac{v}{1-a}\right)^{(1-a)} 
                \geq |w|, u \geq 0, v \geq 0 \right\} 
\end{array}
$$

#### Primal power cone
$$
\begin{array}
\text{minimize}   & x + y \\
\text{subject to} & \sqrt{x} * \sqrt{y} \geq z \\
& x, y \geq 0, z = 4
\end{array}
$$

```{r}
powp <- OP(objective = L_objective(c(1, 1, 0), names = c("x", "y", "z")), 
           constraints = C_constraint(diag(-1, 3), cones = K_powp(0.5), rhs = rep(0, 3)),
           bounds = V_bound(li = 3, ui = 3, lb = 4, ub = 4))
(sol <- ROI_solve(powp))
solution(sol)
```

#### Dual power cone

$$
\begin{array}
\text{minimize}   & u + v \\
\text{subject to} & \sqrt{2 u} * \sqrt{2 v} \geq 4 \\
& u, v \geq 0, z = 4
\end{array}
$$

```{r}
powd <- OP(objective = L_objective(c(1, 1, 0), names = c("x", "y", "z")), 
           constraints = C_constraint(diag(-1, 3), cones = K_powd(0.5), rhs = rep(0, 3)),
           bounds = V_bound(li = 3, ui = 3, lb = 4, ub = 4))
(sol <- ROI_solve(powd))
solution(sol)
```


## Nonlinear Programming
The quadratic problem from above can also be solved by a general purpose solver.

```{r}
(sol <- ROI_solve(qcqp, solver = "alabama", start = double(3)))
solution(sol)
```

or equivalently

```{r}
f <- function(x) {
    90 * x[1] + 110 * x[2] + 160 * x[3] - 1 / 2 * (x[1]^2 + x[2]^2 + x[3]^2)
}
g <- function(x) {
    c((x[1]^2 + x[2]^2 + 4 * x[3]),
      (x[2]^2 + x[3]^2 + x[1] + x[3]),
      (x[1]^2 + x[3]^2 + 2 * x[1] * x[3]))
}
nlp <- OP(F_objective(f, n=3), 
          F_constraint(g, dir = rep("<=", 3), rhs = c(4, 3, 2)),
          maximum = TRUE)
(sol <- ROI_solve(nlp, solver = "alabama", start = double(3)))
solution(sol)
```

or equivalently

```{r}
nlp <- OP(Q_objective(-diag(3), c(90, 110, 160), c("x", "y", "z")), 
          F_constraint(g, dir = rep("<=", 3), rhs = c(4, 3, 2)),
          maximum = TRUE)
(sol <- ROI_solve(nlp, solver = "alabama", start = double(3)))
solution(sol)
```

or equivalently

```{r}
nlp <- OP(F_objective(f, n=3), 
          Q_constraint(Q = list(rbind(c(2, 0, 0), c(0, 2, 0), c(0, 0, 0)),
                                rbind(c(0, 0, 0), c(0, 2, 0), c(0, 0, 2)),
                                rbind(c(2, 0, 2), c(0, 0, 0), c(2, 0, 2))),
                       L = rbind(c(0, 0, 4), c(1, 0, 1), c(0, 0, 0)),
                       dir = rep("<=", 3), rhs = c(4, 3, 2)),
          maximum = TRUE)
(sol <- ROI_solve(nlp, solver = "alabama", start = double(3)))
solution(sol)
```

# Read / Writing optimization problems
Optimization problems are commonly stored and shared in specialized 
plain text files. For LP and MIP the file formats `'mps'`, `'lp'`,
`'sif'` and `'freemps'` are commonly used. 

## Read LP / MILP

```{r}
ROI_registered_reader()
```

To read the capri example from the netlib connection the following code can be used.

```{r}
tmpfile <- tempfile()
con <- gzcon(url("http://www.zib.de/koch/perplex/data/netlib/mps/capri.mps.gz"))
writeLines(readLines(con), tmpfile)
close(con)
(capri <- ROI_read(tmpfile, type="mps_fixed", "lpsolve"))
(sol <- ROI_solve(capri))
```

## Write LP / MILP
```{r}
ROI_registered_writer()
```

```{r}
tmpfile <- tempfile()
ROI_write(milp, tmpfile, type = "lp_lpsolve")
cat(readLines(tmpfile), sep="\n")
```


# Test collections
## NETLIB-LP
The software repository `NETLIB` contains among many other software also
a linear programming test collection.

```{r}
library("ROI.models.netlib")
netlib()
problem_name <- "boeing1"
netlib("metainfo")[problem_name, -(2:5)]
(model <- netlib(problem_name))
(x <- ROI_solve(model))
```

## MILP
```{r}
library("ROI.models.miplib")
miplib("air04")
```

# Options


