---
title: "Ridge regression"
---

[Ridge regression (also known as Tikhonov regularization)](https://en.wikipedia.org/wiki/Tikhonov_regularization) 
shrinks the regression coefficients by adding a quadratic penalty term to the optimization problem.
$$
\underset{\beta}{\text{minimize}} ~~ \frac{1}{2} || y - X \beta ||_2^2 +  \lambda ||\beta||_2^2
$$
with $X \in \mathbb{R}^{m \times n}$, $y \in \mathbb{R}^{m}$, $\beta \in \mathbb{R}^n$
and $0 < \lambda \in \mathbb{R}$.

It is well known that $\beta$ can be estimated by the following formula.
$$
\hat{\beta} = (X^\top X + \lambda I)^{-1} X^\top y
$$
here $I \in R^{n, n}$ refers to the identity matrix.

The optimization problem can be rewritten into a quadratic optimization problem
$$
\begin{array}{rl}
\underset{(\beta, \gamma)}{\text{minimize}} &
    \frac{1}{2} \gamma^\top \gamma + \lambda \beta^\top \beta \\
\text{subject to} & 
    y - X \beta = \gamma
\end{array}
$$
with $\gamma \in \mathbb{R}^n$.


</br>
In **R** the packages [**glmnet**](https://cran.r-project.org/package=glmnet)
and [**MASS**](https://cran.r-project.org/package=MASS) provide functionality 
for ridge regression.


```{r}
y <- longley[,1]
x <- as.matrix(longley[,-1])
```

## Estimation
```{r, message = FALSE}
library(slam)
Sys.setenv(ROI_LOAD_PLUGINS = FALSE)
library(ROI)
library(ROI.plugin.qpoases)

dbind <- function(...) {
    .dbind <- function(x, y) {
        A <- simple_triplet_zero_matrix(NROW(x), NCOL(y))
        B <- simple_triplet_zero_matrix(NROW(y), NCOL(x))
        rbind(cbind(x, A), cbind(B, y))
    }
    Reduce(.dbind, list(...))
}
```

```{r}
qp_ridge <- function(x, y, lambda) {
    stdm <- simple_triplet_diag_matrix
    m <- NROW(x); n <- NCOL(x)
    Q0 <- dbind(stdm(2 * lambda, n), stdm(1, m))
    a0 <- c(b = double(n), g = double(m))
    op <- OP(objective = Q_objective(Q = Q0, L = a0))
    A1 <- cbind(x, stdm(1, m))
    constraints(op) <- L_constraint(A1, eq(m), y)
    bounds(op) <- V_bound(ld = -Inf, nobj = ncol(Q0))
    op
}

op <- qp_ridge(x, y, 0)
(qp0 <- ROI_solve(op, "qpoases"))
cbind(round(coef(lm.fit(x, y)), 3), round(head(solution(qp0), ncol(x)), 3))

op <- qp_ridge(x, y, 10)
(qp1 <- ROI_solve(op, "qpoases"))
head(solution(qp1), NCOL(x))
```

