---
title: "Sudoku"
output:
  html_document:
    self_contained: true
---

It is well known that the number-placement puzzle [Sudoku](https://en.wikipedia.org/wiki/Sudoku)
can be formulated as a binary optimization problem.
The following optimization problem formulation is from [Bartlett et al. (2008)](#BARTLETT).

$$
\begin{array}{rl}
    \underset{x}{\text{minimize}} & \mathbb{0}^\top x \\
    \text{subject to}                          & x_{ijk} \in \{0, 1\} \\
 \text{only one k in each column:}             & \sum_{i=1}^9 x_{ijk} = 1, ~~~~ j = 1:9,~ k = 1:9 \\
 \text{only one k in each row:}                & \sum_{j=1}^9 x_{ijk} = 1, ~~~~ i = 1:9,~ k = 1:9 \\
 \text{only one k in each block:}              & \sum_{j = 3 (q - 1) + 1}^{3 q} \sum_{i = 3(p - 1) + 1}^{3 p} x_{ijk} = 1, ~~~~ k = 1:9,~ p = 1:3,~ q = 1:3 \\
 \text{exactly one number for each position:}  & \sum_{k=1}^9 x_{ijk} = 1, ~~~~ i = 1:9,~ j = 1:9 \\
 \text{stating values are fixed:}              & x_{ijk} = 1 ~~~ \forall (i, j, k) \\
\end{array}
$$

In **R** the packages [**sudoku**](https://CRAN.R-project.org/package=sudoku)
and [**sudokuAlt**](https://CRAN.R-project.org/package=sudokuAlt)
can be used to generate and solve Sudoku puzzles.

```{r, load_sudoku_pkg}
library(sudokuAlt)
sudoku_puzzle <- makeGame()
plot(sudoku_puzzle)
```

```{r, load_roi, message = FALSE}
library(slam)
library(ROI)
library(ROI.plugin.msbinlp)
library(ROI.plugin.glpk)
```

## Auxiliary functions

```{r, auxiliary_functions}
as.matrix.sudoku <- function(x) matrix(as.numeric(x), 9, 9)

to_col_index <- function(i, j, v) {
    (v - 1) * 81 + (i - 1) * 9 + j
}

index_to_triplet <- function(idx) {
    .index_to_triplet <- function(idx) {
        v <- (idx - 1) %/% 81 + 1
        idx <- idx - (v - 1) * 81
        i <- (idx - 1)%/% 9 + 1
        idx <- idx - (i - 1) * 9
        c(i = i, j = idx, v = v)
    }
    t(sapply(idx, .index_to_triplet))
}

solve_sudoku <- function(M, solver, solve = TRUE, ...) {
    stm <- simple_triplet_matrix
    seq9_9 <- rep(1:9, 9)
    seq9_9e <- rep(1:9, each = 9)
    seq81_9e <- rep(seq_len(81), each = 9)
    ones729 <- rep.int(1, 9^3)
    M <- as.matrix(M)
    M[is.na(M)] <- 0
    M <- as.simple_triplet_matrix(M)
    ## setup OP
    op <- OP(double(9^3))
    ## basic setting (fixed coefficients)
    j <- mapply(to_col_index, M$i, M$j, M$v)
    nfv <- length(M$i) ## number of fixed variables
    A0 <- stm(i = seq_len(nfv), j = j, v = rep.int(1, nfv), ncol = 9^3)
    LC0 <- L_constraint(A0, eq(nfv), rep.int(1, nfv))
    ## sum_{i=1:n} x_ijk = 1,  j = 1:n, k = 1:n
    only_one_k_in_each_column <- function(j, k) {
        sapply(1:9, function(i) to_col_index(i, j, k))
    }
    j <- unlist(mapply(only_one_k_in_each_column, seq9_9e, seq9_9, SIMPLIFY = FALSE))
    A1 <- stm(i = seq81_9e, j = j, v = ones729, nrow = 81, ncol = 9^3)
    ## sum_{j=1:n} x_ijk = 1
    only_one_k_in_each_row <- function(i, k) {
        sapply(1:9, function(j) to_col_index(i, j, k))
    }
    j <- unlist(mapply(only_one_k_in_each_row, seq9_9e, seq9_9, SIMPLIFY = FALSE))
    A2 <- stm(i = seq81_9e, j = j, v = ones729, nrow = 81, ncol = 9^3)
    only_one_k_in_each_submatrix <- function(blocki, blockj, k) { 
        i <- (blocki - 1) * 3 + 1:3
        j <- (blockj - 1) * 3 + 1:3
        coo <- expand.grid(i = i, j = j, v = k)
        mapply(to_col_index, i = coo$i, j = coo$j, v = coo$v)
    }
    coo <- expand.grid(i = 1:3, j = 1:3, k = 1:9)
    j <- unlist(mapply(only_one_k_in_each_submatrix, 
                       blocki = coo$i, blockj = coo$j, k  = coo$k, SIMPLIFY = FALSE))
    A3 <- stm(i = seq81_9e, j = j, v = ones729, ncol = 9^3)
    ## at every position in the matrix must be one value
    fill_matrix <- function(i, j) {
        sapply(1:9, function(k) to_col_index(i, j, k))
    }
    j <- unlist(mapply(fill_matrix, i = seq9_9e, j = seq9_9, SIMPLIFY = FALSE))
    A4 <- stm(i = seq81_9e, j = j, v = ones729, ncol = 9^3)
    A <- rbind(A1, A2, A3, A4)
    LC1 <- L_constraint(A, eq(nrow(A)), rep.int(1, nrow(A)))
    constraints(op) <- rbind(LC0, LC1)
    types(op) <- rep.int("B", 9^3)
    if (!solve) return(op)
    s <- ROI_solve(op, solver = solver, ...)
    sol <- solution(s)
    to_sudoku_solution <- function(sol) {
        coo <- index_to_triplet(which(as.logical(sol)))
        sudoku_solution <- as.matrix(stm(coo[,1], coo[,2], coo[,3]), nrow = 9, ncol = 9)
        structure(sudoku_solution, class = c("sudoku", "matrix"))
    }
    
    if ( any(lengths(sol) > 1L) & length(sol) > 1L ) {
        lapply(solution(s), to_sudoku_solution)
    } else {
        if ( length(sol) == 1L ) {
            to_sudoku_solution(sol[[1L]])
        } else {
            to_sudoku_solution(sol)
        }
    }
}

sudoku_is_valid_solution <- function(x) {
    .sudoku_is_valid_solution <- function(x) {
        stopifnot(inherits(x, "sudoku"))
        seq19 <- seq_len(9)
        for (i in seq19) {
            if ( any(sort(as.vector(x[i, ])) != seq19) ) return(FALSE)
            if ( any(sort(as.vector(x[, i])) != seq19) ) return(FALSE)
        }
        for (i in 1:3) {
            for (j in 1:3) {
                block <- x[(i-1) * 3 + 1:3, (j-1) * 3 + 1:3]
                if ( any(sort(as.vector(block)) != seq19) ) return(FALSE)
            }
        }
        return(TRUE)
    }
    if ( is.list(x) ) {
        sapply(x, .sudoku_is_valid_solution)
    } else {
        .sudoku_is_valid_solution(x)
    }
}
```

## Solve Sudoku

We now solve the previously defined Sudoku.

```{r, solve_sudoku}
sudoku_solution <- solve_sudoku(sudoku_puzzle, solver = "glpk")
sudoku_is_valid_solution(sudoku_solution)
plot(sudoku_solution)
```

Furthermore it is easy to check if there exist multiple solutions to the Sudoku.
Here we only want to obtain up to `10` solution to obtain all set
the maximum number of solutions to infinity (`nsol_max = Inf`).

```{r, solve_sudoku_multible_solution}
sudoku_solution <- solve_sudoku(sudoku_puzzle, solver = "msbinlp", nsol_max = 10L)
sudoku_solution
```

### Solve a Sudoku from `http://www.sudoku.org.uk/DailySudoku.asp`
```{r, solve_sudoku_uk}
sudoku_puzzle <- fetchUKGame()
```

Look if this Sudoku has a unique solution.

```{r, sudoku_uk_unique_solution}
sudoku_solution <- solve_sudoku(sudoku_puzzle, solver = "msbinlp", nsol_max = 2L)
has_unique_solution <- inherits(sudoku_solution, "sudoku")
has_unique_solution
```

# References
* Bartlett, A., Chartier, T. P., Langville, A. N., & Rankin, T. D. (2008). An integer programming model for the Sudoku problem. Journal of Online Mathematics and its Applications, 8(1). `URL` [https://www.semanticscholar.org/paper/An-Integer-Programming-Model-for-the-Sudoku-Problem-Bartlett-Chartier/127ccaf7ebde0c47b36ae78cd1e2233b6061a57f](https://www.semanticscholar.org/paper/An-Integer-Programming-Model-for-the-Sudoku-Problem-Bartlett-Chartier/127ccaf7ebde0c47b36ae78cd1e2233b6061a57f) <a name = "BARTLETT"></a>

