---
title: "Poisson regression"
---

To model count data and contingency tables often 
[Poisson regression](https://en.wikipedia.org/wiki/Poisson_regression) is used.
Poisson regression models belong to the 
[generalized linear models](https://en.wikipedia.org/wiki/Generalized_linear_model) 
family (GLM).

Since GLMs are commonly used **R** has already built-in functionality to
estimate GLMs. Specifically the `glm` function from the **stats** package,
withpoisson family and log link can be used to estimate a Poisson model.

The following poisson regression example is from the 
[`glm` manual page](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/glm.html)
and based on [Dobson (1990)](#DOBSON).

```{r, use_case_poisson_regression_glm}
options(width = 10000)
counts <- c(18, 17, 15, 20, 10, 20, 25, 13, 12)
outcome <- gl(3, 1, 9)
treatment <- gl(3, 3)
glm.D93 <- glm(counts ~ outcome + treatment, family = poisson(link = "log"))
round(coef(glm.D93), 4)
```

Making use of maximum likelihood estimation the logistic regression model can also
be estimated in **ROI**. Here either a conic solver or a general purpose solver
can be used. The conic solvers have the advantages that they are specifically 
designed to find the global optimum and are (often) faster.

# Log-likelihood
The maximum likelihood estiamte can be obtained be solving the following
optimzation problem.
$$
\begin{equation}
    \underset{\beta}{\text{maximize}} ~~
      \sum_{i = 1}^n y_i ~ log(\lambda_i) - \lambda_i
      ~~ \text{where} ~~ \lambda_i = exp(X_{i*} \beta)
\end{equation}
$$

# Estimation


```{r, use_case_poisson_regression_data, message = FALSE}
Sys.setenv(ROI_LOAD_PLUGINS = FALSE)
library(ROI)
library(ROI.plugin.nloptr)
library(ROI.plugin.ecos)
X <- model.matrix(glm.D93)
y <- counts
```

# General purpose solver

```{r, use_case_poisson_regression_GPS}
log_likelihood <- function(beta) {
    xb <- drop(X %*% beta)
    sum(y * xb - exp(xb))
}

op_gps <- OP(F_objective(log_likelihood, n =  ncol(X)), maximum = TRUE,
    bounds = V_bound(ld = -Inf, nobj = ncol(X)))
s1 <- ROI_solve(op_gps, "nloptr.lbfgs", start = rnorm(ncol(X)))
round(solution(s1), 4)
```


# Conic solver

This problem can also be estimated by making use of conic optimization.


```{r, use_case_poisson_regression_conic}
library(slam)
poisson_regression <- function(y, X) {
    m <- nrow(X); n <- ncol(X)
    i <- 3 * seq_len(m) - 2
    op <- OP(c(-(y %*% X), rep.int(1, m)))
    stm <- simple_triplet_matrix
    A <- cbind(stm(rep(i, n), rep(seq_len(n), each = m), -drop(X), 3 * m, n),
               stm(i + 2, seq_len(m), rep.int(-1, m), 3 * m, m))
    rhs <- rep(c(0, 1, 0), m)
    cones <- K_expp(m)
    constraints(op) <- C_constraint(A, cones, rhs)  
    bounds(op) <- V_bound(ld = -Inf, nobj = ncol(A))
    op
}

op <- poisson_regression(y, X)
s <- ROI_solve(op, solver = "ecos")
round(head(solution(s), n = NCOL(X)), 4)
```

# References
* Dobson, A. J. (1990) An Introduction to Generalized Linear Models. London: Chapman and Hall. <a name = "DOBSON"></a>
