---
title: "Portfolio Optimization in ROI"
author: 
- "Laura Vana 
- [Email](mailto:laura.vana@wu.ac.at)"
date: "December 4, 2018"
output:
  html_document:
    df_print: paged
---

# Introduction

The purpose of this vignette is to demonstrate a sample of portfolio optimization problems that can be solved by using the ROI package. 
This vignette is based on joint work with Florian Schwendinger and Ronald Hochreiter which was presented at RFinance 2016, Chicago, USA, May 2016.

If previously no ROI version was installed, one should at least install the two plugins `ROI.plugin.glpk`, `ROI.plugin.alabama` and `ROI.plugin.quadprog` to run the examples below.

```{r, message=F}
## NOTE: (Rglpk needs libglpk-dev on Debian/Ubuntu ("sudo apt-get install libglpk-dev"))

if (!require("ROI")) install.packages("ROI"); library("ROI")

if (!require("ROI.plugin.glpk")) install.packages("ROI.plugin.glpk"); library("ROI.plugin.glpk")

if (!require("ROI.plugin.quadprog")) install.packages("ROI.plugin.quadprog"); library("ROI.plugin.quadprog")

if (!require("ROI.plugin.alabama")) install.packages("ROI.plugin.alabama"); library("ROI.plugin.alabama")
```

Several R functions are created to implement the typical objectives and constraints used for portfolio optimization.
All functions require a `data.frame` `r_mat` of returns. The mathematical formulation of the objectives and constraints is presented below.
The default optimization in ROI is minimization. Hence for all maximization problems the argument `maximum = TRUE` in the `OP()` function must be specified.

<br>

### [Objectives](#objectives)  implemented as R functions

Name                                        | Arguments                      |   Objectives                       | Max/Min | Type
--------------------------------------------| -------------------------------| ---------------------------------- | --------|------
[`reward_objective`           ](#rewardo)   | `(r_mat)                     ` | Portfolio return/reward            | max     | LP
[ `markowitz_objective`       ](#markowitzo)| `(r_mat)                     ` | Variance of portfolio              | min     | QP
[`mad_objective          `    ](#mado)      | `(r_mat)                     ` | Mean Absolute Deviation            | min     | LP
[`downside_var_objective `    ](#downvaro)  | `(r_mat)                     ` | Lower semi-variance of portfolio   | min     | QP
[`downside_mad_objective `    ](#downmado)  | `(r_mat)                     ` | Lower semi-mean absolute deviation | min     | LP
[`cvar_objective         `    ](#cvaro)     | `(r_mat, alpha, probs = NULL)` | Conditional Value-at-Risk          | min     | LP
[`minimax_young_objective`    ](#minimaxo)  | `(r_mat)                     ` | Minimax portfolio                  | max     | LP 
[`quadratic_utility_objective`](#quaduo)    | `(r_mat, lambda = 2)            `          | Quadratic Utility                  | max     | LP 
[`sharpe_objective       `    ](#sharpeo)   | `(r_mat, rf = 0)             ` | Sharpe ratio                       | max     | QP 
[`omega_objective        `    ](#omegao)    | `(r_mat, tau = 0)            ` | Omega                              | max     | LP 

<br>

### [Constraints](#constraints) implemented as R functions

Name            | Arguments                              |   Constraints                                                        | Type
--------------- | -------------------------------------- | -------------------------------------------------------------------- | ----
[`budget_constraint`](#budgetc) | `(r_mat, dir = "==", rhs = 1)`| Budget constraint,i.e., sum of the portfolio equals budget B (default B=`rhs`=1)| LP
[`group_constraint` ](#groupc)  | `(r_mat, index, coef.index = 1, dir = "==", rhs) `   | Group constraint applied only the the `index` elements of $x$       | LP 
[`turnover_constraint   ` ](#turnoverc)    | `(r_mat, x0 = NULL, dir = "<=", rhs = 100) ` | Constraint on the turnover  from some initial weights $x_0$   | LP 
[`cardinality_constraint`](#cardc)  | `(r_mat, dir = "<=", rhs = 100)` | Cardinality            | MILP
[`reward_constraint     `](#rewardc) | `(r_mat, dir = ">=", rhs = 0)             ` | Target return constraint                                                    | LP
[`markowitz_constraint  `](#markc)   | `(r_mat, dir = "<=", rhs = 1000)` | Constraint on variance of portfolio $x^{\top}Qx \leq q^2$                                               | QP
[`cvar_constraint       ` ](#cvarc)  | `(r_mat, alpha, probs = NULL, dir = "<=", rhs = 100)` | Constraints on the conditional Value-at-Risk                                            | LP



# Objective functions {#objectives} 
We consider a portfolio of assets with random returns. We denote the portfolio
choice vector by $x$ and by $r$ the vector of random returns. 
$N$ denotes the number of assets in the portfolio while $S$ is the number of scenarios in the scenarios set.

## Maximize expected return {#rewardo}
$$ \max_{x \in \mathbb{R}^N} \hat \mu^\top x $$
where $\hat \mu$ is the vector of estimated mean asset returns.

In ROI we define a function which returns a linear objective is `L_objective` (note that the minus turns the maximization into a minimization problem):

```{r}
reward_objective <- function(r_mat) {
    objective <- L_objective(colMeans(r_mat))
    list(objective = objective)
}
```
(see [Example 1](#rewardex1), [Example 12](#rewardex2), [Example 13](#rewardex3)).

## Minimize variance {#markowitzo} 
The objective of the minimum of the variance portfolio is given by:
\[\min_{x \in \mathbb{R}^N} x^\top Q x\]
\[Q = \mathbb{C}ov(r)\]

```{r}
markowitz_objective <- function(r_mat) {
    objective <- Q_objective(Q = 2 * cov(r_mat), 
                             L = rep(0, NCOL(r_mat)))
    list(objective = objective)
}
```
(see [Example 2](#markex1), [Example 3](#markex2)).

## Minimize mean absolute deviation {#mado}
The minimization of the mean absolute deviation can be written as a linear problem. 
$$
\begin{eqnarray*}
\min_{x\in \mathbb{R}^N, y\in \mathbb{R}^S, z\in \mathbb{R}^S}&& \frac{1}{S} \sum_{s = 1}^S (y_s + z_s)\\
s.t.&&y_s - z_s = (r_{s} - \hat \mu)^\top x,\\
&& y_s \geq 0, z_s \geq 0.
\end{eqnarray*}
$$
Note that the bounds $y_s\geq 0$ and $z_s\geq 0$ must not be additionally specified as these are the default in ROI.

```{r}
mad_objective <- function(r_mat){
    x.names <- colnames(r_mat)
    N <- NCOL(r_mat)
    S <- nrow(r_mat)
    mu <- colMeans(r_mat)
    Amat <-  cbind(sweep(as.matrix(r_mat), 2,  mu), - diag(S), diag(S))
    var.names <- c(x.names, 
                   paste0("y_mad_aux", seq_len(S)),
                   paste0("z_mad_aux", seq_len(S)))
    
    constraint <-  L_constraint(L = Amat, dir = rep("==", S), 
                                rhs = rep(0, S), 
                                names = var.names)

    objective <- L_objective(L = c(rep(0, N), rep(1/S, 2 * S)))
    
    list(objective = objective, constraint = constraint)
}
```
(See [Example 4](#madex1))

## Minimize lower semi-variance {#downvaro}
A downside risk measure is the lower semi-variance, which is the expected squared deviation from the mean, 
calculated over those points that are no greater than the mean. 
Only returns that are below the mean 
contribute to the portfolio risk. Let $r_s$ denote the vector of random returns of the $N$ assets in scenario $s$. 
The problem can be formulated as (using the auxiliary variable $z\in \mathbb{R}^S$):
$$\min_{x\in \mathbb{R}^N} \frac{1}{S} \sum_{s = 1}^S \mathrm{max}(\hat\mu^\top x - r_s^\top x, 0)^2.$$
The problem can be reformulated as a QP:
$$
\begin{eqnarray*}
\min_{x\in \mathbb{R}^N, z\in \mathbb{R}^S}&& \frac{1}{S} \sum_{s = 1}^S z_s^2\\
s.t.&&z_s \geq (\hat\mu - r_s)^\top x)\\
&& z_s \geq 0.
\end{eqnarray*}
$$

```{r}
downside_var_objective <- function(r_mat, x.names = NULL) {
    x.names <- colnames(r_mat)
    N <- NCOL(r_mat)
    S <- NROW(r_mat)
    mu <- colMeans(r_mat)
    Amat <- cbind(sweep(as.matrix(r_mat), 2, mu), diag(S))
    var.names <- c(x.names, paste0("z_dvar_aux", seq_len(S)))
    
    constraint <- L_constraint(L = Amat, dir = rep(">=", S), 
                               rhs = rep(0, S), 
                               names = var.names)
    objective <- Q_objective(
             Q = 2 * diag(c(rep(1e-05, N), rep(1/S, S))),
             L = rep(0, N + S))
    
    list(objective = objective, constraint = constraint)
}
```
(see [Example 5](#downvarex1)).

## Minimize lower semi-absolute deviation {#downmado}
Let $r_s$ denote the vector of random returns of the $N$ assets in scenario $s$. The downside risk measure given by the expected absolute deviation from the mean,  calculated over those points that are no greater than the mean is given by:
$$\frac{1}{S} \sum_{s = 1}^S \vert \mathrm{max}(\hat\mu^\top x - r_s^\top x, 0) \vert$$
The minimization of this risk measure can be reformulated as a linear problem:
$$
\begin{eqnarray*}
\min_{x\in \mathbb{R}^N, z\in \mathbb{R}^S}&& \frac{1}{S} \sum_{s = 1}^S z_s,\\
s.t.&& z_s \geq (\hat\mu - r_s)^\top x\\
&& z_s \geq 0.
\end{eqnarray*}
$$

```{r}
downside_mad_objective <- function(r_mat){
    x.names <- colnames(r_mat)
    N <- NCOL(r_mat)
    S <- NROW(r_mat)
    mu <- colMeans(r_mat)
    Amat <- cbind(sweep(as.matrix(r_mat), 2, mu), diag(S))
    var.names <- c(x.names, paste0("z_dmad_aux", seq_len(S)))
    
    constraint <- L_constraint(L = Amat, dir = rep(">=", S),
                               rhs = rep(0, S), 
                               names = var.names)
    
    objective  <- L_objective(L = c(rep(0, N), rep(1/S,  S)))
    
    list(objective = objective, constraint = constraint)    
}
```
(see [Example 6](#downmadex1)).

## Minimize conditional value at risk {#cvaro}

The conditional value at risk/expected shortfall is the average of the losses  that exceed the $\alpha$-quantile of the loss distribution, also called $\alpha$ Value-at-Risk (VaR).
Defining the loss distribution as minus of the return of the portfolio, the problem can be formulated as follows:
$$
\begin{eqnarray*}
\min_{x\in \mathbb{R}^N, z\in \mathbb{R}^S, \gamma\in \mathbb{R}}& \gamma + \frac{1}{(1-\alpha)} \sum_{s = 1}^S p_s z_s\\
s.t.\ &z_s \geq - r_s^\top x - \gamma,\\
&z_s \geq 0,
\end{eqnarray*}
$$
where VaR is a minimizer of the function over $\gamma$ and $p=(p_1, \dots, p_S)$ is a vector of probabilities associated with the scenarios.

```{r}
cvar_objective <- function(r_mat, alpha, probs = NULL) {
  x.names <- colnames(r_mat)
  N <- NCOL(r_mat)
  S <- NROW(r_mat)
  mu <- colMeans(r_mat)
  if (is.null(probs)) probs <- rep(1/S, S)
  if (alpha < 0.5) alpha <- 1 - alpha
  
  Amat <- cbind(as.matrix(r_mat),  diag(S), 1)
  var.names <- c(x.names, paste0("z_cvar_aux", seq_len(S)), "gamma")
  
  ## set bounds for gamma (-Inf, Inf) 
  bnds <- ROI::V_bound(li = c(N + S + 1), lb = c( -Inf),
                       ui = c(N + S + 1), ub = c(  Inf))
  
  constraint <- L_constraint(L = Amat, dir = rep(">=", S), 
                            rhs = rep(0, S), 
                            names = var.names)

  objective <- L_objective(c(rep(0, N), probs/(1 - alpha), 1))

  list(objective = objective, constraint = constraint, bounds = bnds)
}
```
Note that if no value is specified for $p_1, \cdots, p_S$ the default is equal weights.

(see [Example 7](#cvarex1), [Example 14](#cvarex3)).

## Minimax Portfolio {#minimaxo}
The optimal portfolio is defined as that one that minimizes the maximum loss over all past historical periods, subject to a restriction on the minimum acceptable average return across all observed periods of time. 
The minimax portfolio maximizes the minimum gain
and can be seen as a limiting case of CVaR for $\alpha \rightarrow 1$.
Let $M_p=\min_s r_s^\top x$ denote the minimum gain of the portfolio.
$$
\begin{eqnarray*}
\max_{x \in \mathbb{R}^N, M_p\in \mathbb{R}}&&  M_p\\
s.t.&% r_s^\top x - M_p \geq 0.
\end{eqnarray*}
$$
Young, M.R., 1998. A minimax portfolio selection rule with linear programming solution. _Management science_, 44(5), pp.673-683.

```{r}
minimax_young_objective <- function(r_mat){
    x.names <- colnames(r_mat)
    N <- NCOL(r_mat)
    S <- NROW(r_mat)
    Amat <- cbind(as.matrix(r_mat), -1)
    
    bnds <- ROI::V_bound(li = c(N + 1), lb = c( -Inf),
                          ui = c(N + 1), ub = c(  Inf))
    constraint <- L_constraint(L = Amat, dir=rep(">=", S), rhs=rep(0, S), 
                                      names=c(x.names,  "mp_aux"))
  
    objective <- L_objective(c(rep(0, N), 1))
  
    list(objective = objective, constraint = constraint, bounds = bnds)     
}
```
(see [Example 8](#minimaxex1)).

## Maximize quadratic utility {#quaduo}
Let $\lambda$ denote the risk aversion parameter.
Typical risk aversion parameters lie between 2 and 4. 
Then the quadratic utility is given by:
\begin{eqnarray*}
\max_{x\in\mathbb{R}^N}&&  \hat \mu^\top x - \frac{\lambda}{2} x^\top Q x
\end{eqnarray*}

```{r}
quadratic_utility_objective <- function(r_mat, lambda = 2){
    objective <- Q_objective(Q = - lambda * cov(r_mat), 
                             L = colMeans(r_mat))
    list(objective = objective)
}
```
(see [Example 9](#quaduex1)).

## Maximize Sharpe ratio {#sharpeo}

Let $r_f$ denote the risk free rate (when not specified is set by default to zero).
The Sharpe ratio of a random return of an asset $\tilde r$  is defined as:
$$ \frac{\mathbb{E}(\tilde r)- r_f}{\sqrt{\mathbb{V}ar(\tilde r)}} $$
The budget normalization constraint is applied and the problem can be formulated as a QP:
$$
\begin{eqnarray*}
\max_{y\in \mathbb{R}^N, \kappa\in \mathbb{R}}& -  y^\top Q y\\
s.t.&(\hat\mu - r_f \mathit{1})^\top y = \text{1},\\
& \mathit{1}^\top y = \kappa,\\
& \kappa > 0.
\end{eqnarray*}
$$
The optimal solution is given by $x^* = y^* / \kappa^*$. 

```{r}
sharpe_objective <- function(r_mat, rf = 0){
    N <- NCOL(r_mat)
    S <- NROW(r_mat)
    mu <- colMeans(r_mat)
    Amat <- rbind(c(mu - rf, 0),
                  c(rep(0, N), 1),
                  c(rep(1, N), -1))
    var.names <- c(paste0("y_sharpe_aux", seq_len(N)), "kappa_sharpe")
    
    constraint <- L_constraint(L = Amat, dir = c("==", ">", "=="),
                               rhs = c(1, 0, 0), names = var.names)
    
    mat <- matrix(0, ncol = N + 1, nrow = N + 1)
    mat[1:N, 1:N] <- 2 * cov(r_mat)
    mat[N + 1, N + 1] <-  1e-04

    objective <- Q_objective(Q = - mat, L = c(rep(0, N), 0))
  
    list(objective = objective, constraint = constraint)     
}
```
(see [Example 10](#sharpeex1)).

## Maximize Omega {#omegao}
The Omega Ratio is a risk-return performance measure of an investment asset, portfolio, or strategy. 
The Omega Ratio, introduced in 2002 by Keating and Shadwick, is defined as the probability weighted ratio of gains versus losses for some threshold return target $\tau$.
$$\Omega(\tilde r) = \frac{\int_\tau^{+\infty}\left(1 - F(r)\right)\mathrm{d}r}{\int_{-\infty}^\tau F(r)\mathrm{d}r} $$
where $F(r)$ denotes the cumulative distribution function of $\tilde r$.
The maximization of $\Omega(\tilde r)$ can be formulated as an linear problem (note that the  budget normalization constraint and the target return constraint are already part of the model formulation).
\begin{eqnarray*}
\max_{y\in \mathbb{R}^N, u\in \mathbb{R}^S, z\in \mathbb{R}}&  \hat \mu^\top y - \tau z\\
s.t.\ & u_s \geq \tau - r_s^\top y , \\
             & u_i\geq 0,\\ 
             & \mathit{1}^\top u = 1,\\ 
             & \mathit{1}^\top y = z,\\ 
             & \hat \mu^\top y \geq \tau z,\\
             & z \geq 0  .
\end{eqnarray*}
The Omega optimal portfolio is given by  $x^* = y^* /z^*$, where $z$ is the homogenizing variable. If not explicitly specified, the default for the threshold is $\tau = 0$.

```{r}
omega_objective <- function(r_mat, tau = 0){
  ## variables y_1, ... y_N, u_1, .., u_S, z
  N <- NCOL(r_mat)
  S <- NROW(r_mat)
  mu <- colMeans(r_mat)
  
  Amat <- rbind(cbind(as.matrix(r_mat), diag(S), 0),# u_s >= tau - r_s'y
                c(rep(0, N), rep(1, S),  0), # sum(u) = 1
                c(rep(1, N), rep(0, S), -1), # sum(y) = z
                c(mu,        rep(0, S), - tau),
                c(rep(0, N), rep(0, S),  1)) # mu'y  >= tau * z
  var.names <- c(paste0("y_omega_aux", seq_len(N)),
                 paste0("u_omega_aux", seq_len(S)), 
                 "z_omega")
  constraint <-  L_constraint(L = Amat,  
                              dir = c(rep(">=", S), "==", "==", ">=", ">"),
                              rhs = c(rep(tau, S), 1, 0, 0, 1e-05), 
                              names=var.names)
  objective  <- L_objective(L = c(mu, rep(0, S), -tau))
  
  list(objective = objective, constraint = constraint)    
  ## x* <- y*/z*
}
```
(see [Example 11](#omegaex1)).

# Constraints {#constraints}

##  Box constraints

Can be set by the argument `bounds` in `OP()`. Note that the default in ROI is $0 \leq x_i \leq \infty$. 

## Budget constraint {#budgetc}
$$\sum_{i=1}^N x_i = B \qquad \sum_{i=1}^N x_i \leq B_u \qquad \sum_{i=1}^N x_i \geq B_l.$$
Budget normalization constraint is obtained for B = 1 (default).

```{r}
budget_constraint <- function(r_mat, dir = "==", rhs = 1) {
    x.names <- colnames(r_mat)
    L_constraint(L = rep(1, NCOL(r_mat)), 
                 dir = dir,  rhs = rhs, names = x.names)
}
```
## Group constraints {#groupc}
Group constraints are linear constraints which apply to only some elements of the portfolio choice vector $x$:
$$ ax_i + bx_j = c $$
where the $=$ can be replaced by inequalities.

```{r}
group_constraint <- function(r_mat, index, coef.index = 1, dir = "==", rhs) {
  ## index = (i, j)
  ## coef.index = c(a,b)
  ## rhs = c
  x.names <- colnames(r_mat)
  N <- NCOL(r_mat)
  L <- rep(0, N)
  L[index] <- coef.index
  L_constraint(L = L, dir = dir, rhs = rhs, names = x.names)
}

```
(see [Example 1](#rewardex1), [Example 12](#rewardex2)).

## Turnover constraint {#turnoverc}
For a target turnover $L$ and some initial weights $x_0$, the turnover contraint is $$\sum_{i=1}^N \vert x_i - x_{i0}\vert \leq L.$$ 
This can be reformulated in terms of auxiliary variables
$y^+, y^- \in \mathbb{R}^{+N}$ such that 
\begin{eqnarray*}
&&y^+_i- y_i^- = x_i - x_{0i}\\
&&\sum_{i=1}^N (y_i^+ + y_i^-)\leq L\\
&&y^+ \geq 0, \, y^-\geq 0
\end{eqnarray*}
If no initial weights are specified, the default is equal weights. Note: does not work out of the box with Omega and Sharpe objectives.

```{r}
turnover_constraint <- function(r_mat, x0 = NULL, dir = "<=", rhs = 100) {
    x.names <- colnames(r_mat)
    N <- NCOL(r_mat)
    S <- NROW(r_mat)
    if (is.null(x0)) x0 <- rep(1/N, N)
    Amat <- cbind(diag(N), - diag(N), diag(N))
    var.names <- c(x.names,  
                   paste0("y_plus_aux", seq_len(N)), 
                   paste0("y_minus_aux", seq_len(N)))
    
    rbind(L_constraint(L = Amat, dir = rep("==", N), rhs = x0, 
                       names = var.names),
          L_constraint(c(rep(0, N), rep(1, N), rep(1, N) ), 
                       dir = dir, rhs = rhs, names = var.names))
}
```
(see [Example 12](#rewardex2)).

## Cardinality constraint {#cardc}
The cardinality constraint assigns bounds $P_{min}$ and $P_{max}$ on the number of assets in a portfolio. Binary auxiliary variables $z\in \mathbb{R}^N$ are introduced such that $P_{min} \leq \sum_{i=1}^N z_i\leq P_{max}$ and $x_i - z_i \leq 0$. 

```{r}
cardinality_constraint <- function(r_mat, dir = "<=", rhs = 100) {
  x.names <- colnames(r_mat)
  N <- NCOL(r_mat)
  Amat <- cbind(diag(N), -diag(N))
  var.names <- c(x.names, paste0("z_card_aux", seq_len(N)))
  cat("Variable types for z_card_aux must be set to binary.\n")
  rbind(L_constraint(L = Amat, dir = rep("<=", N), 
                     rhs = rep(0, N), names = var.names), 
        L_constraint(L = c(rep(0, N), rep(1, N)), dir = dir,  
                     rhs = rhs, names = var.names))
}
```
(see [Example 14](#cvarex2)).

## Target return constraint {#rewardc}
Also typical objective functions can be used as constraints. For example, the expected return must not fall below a value $\tau$:
$$\hat\mu^\top x \geq \tau$$
Note: does not work with Sharpe or Omega (it is included in both Sharpe and Omega maximization problems).

```{r}
reward_constraint <- function(r_mat, dir = ">=", rhs = 0) {
  x.names <- colnames(r_mat)
  L_constraint(L = colMeans(r_mat), dir = dir,  
               rhs = rhs, names = x.names)
}
```
(see [Example 3](#markex1), [Example 6](#downmadex1)).

## Risk constraints -- Variance {#markc}
Typically constraints of the form $x^\top Q x \leq q^2$. Note however that this is a quadratic constraint and needs appropriate solvers (e.g., in ROI.plugin.alabama).
Note: does not work with Omega.

```{r}
markowitz_constraint <- function(r_mat, dir = "<", rhs = 1000) {
    x.names <- colnames(r_mat)
    N <- NCOL(r_mat)
    Q_constraint(Q = 2 * cov(r_mat), L = rep(0, N), 
                 dir = dir, rhs = rhs, names=x.names)
}
```
(see [Example 13](#rewardex3)).

## Risk constraints -- CVaR {#cvarc}
Typically CVaR$(x, \alpha) \leq q$. 
Note: does not work with Omega and Sharpe.

```{r}
cvar_constraint <- function(r_mat, alpha, probs = NULL, dir = "<=", rhs = 100) {
  x.names <- colnames(r_mat)
  N <- NCOL(r_mat)
  S <- NROW(r_mat)
  if (alpha < 0.5) alpha <- 1 - alpha
  if (is.null(probs)) probs <- rep(1/S, S)
  
  Amat <- cbind(as.matrix(r_mat), diag(S), 1)
  var.names <- c(x.names, paste0("z_cvar_aux", seq_len(S)), "gamma")
  # set bounds for gama
  bnds <- ROI::V_bound(li = c(N + S + 1), lb = c( -Inf),
                       ui = c(N + S + 1), ub = c(  Inf))
  rbind(L_constraint(L = Amat, dir = rep(">=", S), 
                     rhs=rep(0, S), names = var.names), 
        L_constraint(c(rep(0, N), rep(1 / ((1 - alpha) * S), S), 1), 
                     dir = dir, rhs = rhs, names = var.names))
}    
```
(see [Example 12](#rewardex2)).


# Data 
For illustration purposes daily log returns for the past 180 days of the 30 companies included in the 2018 Dow Jones Industrial Average will be used. 

```{r, results = 'hide', message=FALSE}
if (!require("quantmod")) install.packages("quantmod"); library("quantmod")
Tickers     <- c("MMM", "AXP", "AAPL", "BA", "CAT", "CVX", "CSCO", "KO", "DIS", "DOW",
                 "XOM", "GS", "HD", "IBM", "INTC", "JNJ", "JPM",
                 "MCD", "MRK", "MSFT", "NKE", "PFE", "PG",
                 "TRV", "UNH", "UTX", "VZ", "V", "WMT", "WBA")

cached_file <- "cache/cached_portfolio_data.rda"
if ( file.exists(cached_file) ) {
  load(cached_file)
} else {
  download_tickers <- function(x) {
    download_ticker_error_handler <- function(e) {
      warning(paste(x, 'not found'))
      return(NA)
    }
    tryCatch(getSymbols(x, from = Sys.Date() - 180, auto.assign = FALSE),
             error = download_ticker_error_handler)
  }
  Historicals <- lapply(Tickers, download_tickers)

  prices <- do.call("cbind",lapply(Historicals, function(x) x[, 6]))

  djia2018 <- as.data.frame((log(lag(prices)) - log(prices))[-1, ])
  save(Historicals, prices, djia2018, file = cached_file)
}
```

```{r, eval=FALSE, results = 'hide', message=FALSE}
if (!require("quantmod")) install.packages("quantmod"); library("quantmod")
Tickers     <- c("MMM", "AXP", "AAPL", "BA", "CAT", "CVX", "CSCO", "KO", "DIS", "DOW",
                 "XOM", "GS", "HD", "IBM", "INTC", "JNJ", "JPM",
                 "MCD", "MRK", "MSFT", "NKE", "PFE", "PG",
                 "TRV", "UNH", "UTX", "VZ", "V", "WMT", "WBA")
download_tickers <- function(x) {
  download_ticker_error_handler <- function(e) {
      warning(paste(x, 'not found'))
      return(NA)
    }
  tryCatch(getSymbols(x, from = Sys.Date() - 180, auto.assign = FALSE),
           error = download_ticker_error_handler)
}
Historicals <- lapply(Tickers, download_tickers)

prices <- do.call("cbind",lapply(Historicals, function(x) x[, 6]))

djia2018 <- as.data.frame((log(lag(prices)) - log(prices))[-1, ])
```

We consider a portfolio of assets with random returns. We denote the portfolio
choice vector by $x$ and by $r$ the vector random returns. 
$N$ denotes the number of assets in the portfolio while $S$ is the number of scenarios in the scenarios set.


<!-- ############################################# -->
# Examples 

## Example 1: Maximize expected return subject to budget normalization and group constraints {#rewardex1}
The following optimization problem:
$$
\begin{eqnarray*}
\max_{x \in \mathbb{r}^N}&& \hat \mu^\top x\\
\sum_{x=1}^N x_i &=& 1\\
x_i &\geq& 0\\
x_3 + x_{17} &\leq& 0.5 
\end{eqnarray*} 
$$
can be set up easily in ROI:

```{r}
lp  <- OP(objective  =  reward_objective(djia2018)$objective,
          constraints = rbind(budget_constraint(djia2018),
                              group_constraint(djia2018, index = c(3, 17), dir = "==", rhs = 0.5)),
          maximum = T)
```
(Note that the bounds can be omitted as this is the default in ROI). 
To perfom the optimization, the `ROI_solve()` function is called:
```{r}
(sol <- ROI_solve(lp, solver = "glpk"))
```

```{r}
solution(sol)
```

## Example 2: Minimum variance portfolio {#markex1}

```{r}
lp  <- OP(objective  =  markowitz_objective(djia2018)$objective,
          constraints = rbind(budget_constraint(djia2018)))
```
(Note that the bounds can be omitted as this is the default in ROI). 
To perfom the optimization, the `ROI_solve()` function is called:
```{r}
(sol <- ROI_solve(lp, solver = "quadprog"))
```

```{r}
round(solution(sol), 3)
```
## Example 3: Minimum variance portfolio with budget normalization and target return constraint {#markex2}

```{r}
lp  <- OP(objective  =  markowitz_objective(djia2018)$objective,
          constraints = rbind(budget_constraint(djia2018),
                              reward_constraint(djia2018, rhs = 0.001)))
```
To perfom the optimization, the `ROI_solve()` function is called:

```{r}
(sol <- ROI_solve(lp, solver = "quadprog"))
```

```{r}
round(solution(sol), 3)
```
## Example 4: Minimize mean absolute deviation portfolio {#madex1}
The minimum MAD portfolio is given by:
```{r}
lp  <- OP(objective  =  mad_objective(djia2018)$objective,
          constraints = rbind(mad_objective(djia2018)$constraint,
                              budget_constraint(djia2018),
                              use.names = TRUE))

```
(Note that the bounds can be omitted as this is the default in ROI). 
To perfom the optimization, the `ROI_solve()` function is called:

```{r}
(sol <- ROI_solve(lp, solver = "glpk"))
```
The optimal weights are:

```{r}
round(solution(sol)[seq_len(NCOL(djia2018))], 3)
```

## Example 5: The  minimum lower semi-variance portfolio {#downvarex1}
The  minimum lower semi-variance portfolio is given by:

```{r}
tmp <- downside_var_objective(djia2018)
lp  <- OP(objective  = tmp$objective,
          constraints = rbind(tmp$constraint,
                              budget_constraint(djia2018), 
                              use.names = TRUE))

```
To perfom the optimization, the `ROI_solve()` function is called:

```{r}
(sol <- ROI_solve(lp, solver = "quadprog"))
```
The optimal weights are:

```{r}
round(solution(sol)[seq_len(NCOL(djia2018))], 3)
```

## Example 6: Minimum lower semi-absolute deviation portfolio with target return constraint {#downmadex1}
The portfolio with minimum lower semi-absolute deviation and minimum 0.001 target return is given by:

```{r}
tmp <- downside_mad_objective(djia2018)
lp  <- OP(objective  = tmp$objective,
          constraints = rbind(tmp$constraint,
                              budget_constraint(djia2018),
                              reward_constraint(djia2018, rhs = 0.001),
                              use.names = TRUE))

```
To perfom the optimization, the `ROI_solve()` function is called:

```{r}
(sol <- ROI_solve(lp, solver = "glpk"))
```

```{r}
round(solution(sol)[seq_len(NCOL(djia2018))], 3)
```

## Example 7: Minimum 95\% & 99\% conditional value at risk:
The portfolio with minimum 95\% conditional value at risk can be obtained as follows:

```{r}
tmp <- cvar_objective(djia2018, alpha = 0.95)
lp  <- OP(objective  =  tmp$objective,
          constraints = rbind(tmp$constraint,
                              budget_constraint(djia2018),
                              use.names = TRUE),
          bounds = tmp$bounds)
```
To perfom the optimization, the `ROI_solve()` function is called:

```{r}
(sol <- ROI_solve(lp, solver = "glpk"))
```

```{r}
round(solution(sol)[seq_len(NCOL(djia2018))], 3)
```
The solution for 99\% CVaR is given by:

```{r}
tmp99 <- cvar_objective(djia2018, alpha = 0.99)
lp99  <- OP(objective  =  tmp99$objective,
          constraints = rbind(tmp99$constraint,
                              budget_constraint(djia2018),
                              use.names = TRUE),
          bounds = tmp$bounds)
sol99 <- ROI_solve(lp99, solver = "glpk")
```
```{r}
round(solution(sol99)[seq_len(NCOL(djia2018))], 3)
```
The 95\% or the 99\% VaR are a by-product of the optimization problem and can be extracted from the solution

```{r}
VaR <- c("0.95" = sol$solution["gamma"], 
         "0.99" = sol99$solution["gamma"]) 
VaR
```

## Example 8: The minimax portfolio {#minimaxex1}

```{r}
tmp <- minimax_young_objective(djia2018)
lp  <- OP(objective  =  tmp$objective,
          constraints = rbind(tmp$constraint,
                              budget_constraint(djia2018),
                              use.names = TRUE),
          bounds =  tmp$bounds,
          maximum =  T)

```
To perfom the optimization, the `ROI_solve()` function is called:

```{r}
(sol <- ROI_solve(lp, solver = "glpk"))
```
```{r}
round(solution(sol)[seq_len(NCOL(djia2018))], 3)
```
## Example 9: Maximize quadratic utility where short selling is allowed but the weights should not be less than -1. {#quaduex1}

```{r}
tmp <- quadratic_utility_objective(djia2018)
lp  <- OP(objective  =  tmp$objective,
          constraints = budget_constraint(djia2018),
          bounds = V_bound(li = seq_len(NCOL(djia2018)), 
                           lb = rep(-1, NCOL(djia2018))),
          maximum = T)
```

```{r}
(sol <- ROI_solve(lp, solver = "quadprog"))
```
```{r}
round(solution(sol)[seq_len(NCOL(djia2018))], 3)
```

## Example 10: Maximize Sharpe ratio with shortselling {#sharpeex1}

```{r}
N <- NCOL(djia2018)
tmp <-sharpe_objective(djia2018)

lp <- OP(maximum = T)
objective(lp) <- tmp$objective
```
When imposing constraints on the portfolio choice vector $x$ using the Sharpe objective, one must keep in mind that $x^* = y^* / \kappa^*$, where $y$ and $\kappa$ are the values optimized. 
Hence, the constraint $x_i \geq -1$ must be transformed to:
$$ y_i/\kappa \geq -1 \Rightarrow y_i \geq -\kappa \Rightarrow y_i + \kappa \geq 0 $$:

```{r}
mat <- cbind(diag(N),  1)
shortsell_constraint <- L_constraint(mat, dir = rep(">=", N),
                                     rhs = rep(0, N))

constraints(lp) <- rbind(tmp$constraint,
                         shortsell_constraint)
```
Moreover, the lower bounds for the auxiliary variables must be set to `-Inf`.

```{r}
bounds(lp) <- V_bound(li = seq_len(N +1), 
                      lb = c(rep(-Inf, N), 0))
```

```{r}
(sol <- ROI_solve(lp, solver = "quadprog"))
```
The optimal solution is given by $x^* = y^* / \kappa^*$: 

```{r}
sol_sharpe <- solution(sol)
x_opt <- round(sol_sharpe[1:30]/sol_sharpe["kappa_sharpe"], 3)
names(x_opt) <-colnames(djia2018)
x_opt
```

## Example 11: Maximize Omega {#omegaex1}

```{r}
tmp <- omega_objective(djia2018)
lp  <- OP(objective  =  tmp$objective,
          constraints = tmp$constraint,
          maximum = T)
```

```{r}
(sol <- ROI_solve(lp, solver = "glpk"))
```
The optimal solution is given by $x^* = y^* / z^*$. 
```{r}
sol_omega <- solution(sol)
x_opt <- round(sol_omega[1:30]/sol_omega["z_omega"], 3)
names(x_opt) <-colnames(djia2018)
x_opt
```

## Example 12: Maximize expected returns wrt to cvar, turnover, short selling and group constraints {#rewardex2}

\begin{eqnarray*}
\max_{x\in\mathbb{R}^N}&  \hat\mu^\top x \\
s.t.& \mathrm{CVaR}(x, 0.95) \leq 0.02 \\
             & x_i \geq  -1\\
             & x_2 + x_{10} + x_{20}\leq 0.5 \\
             & \sum_{i=1}^N \vert x_i - x_{0i} \vert \leq 0.5
\end{eqnarray*}
Given that the constraints have more (auxiliary) variables than the objective, first the constraints are defined:

```{r}
lp <- OP(maximum = T)

constraints(lp) <- rbind(
  budget_constraint(djia2018),
  group_constraint(djia2018, index = c(2, 10, 20), dir = "<=", rhs = 0.5),
  turnover_constraint(djia2018, dir = "<=", rhs = 0.5),
  cvar_constraint(djia2018, alpha = 0.95, rhs = 0.02),
  use.names = TRUE)
```
Next, the  vector of the linear objective is filled with zeros to match the dimension of the constraints:

```{r}
obj <- c(terms(reward_objective(djia2018)$objective)$L)
objective(lp) <- c(obj, double(ncol(constraints(lp)) - length(obj)))
(sol <- ROI_solve(lp, solver = "glpk"))
solution(sol)[1:30]
```
## Example 13:  Maximize expected return wrt variance constraint {#rewardex3}
\begin{eqnarray*}
\max_{x\in\mathbb{R}^N}&&  \hat\mu^\top x \\
s.t.&& x^{\top}Qx \leq 0.1 \\
             &&\sum_{i=1}^N x_i =1\\
             && x_i \geq 0
\end{eqnarray*}

```{r}
p <- OP(maximum = T)

constraints(p) <- rbind(
  markowitz_constraint(djia2018, dir = "<=", rhs = 0.5^2),
  budget_constraint(djia2018),
  use.names = TRUE)

objective(p) <- reward_objective(djia2018)$objective 
```
When using the alabama solver starting values need to be chosen. 
A good strategy is to use several starting values and compare the results.

```{r}
mstart <- lapply(1:10, function(x) runif(length(objective(p))))
solus <- lapply(mstart, function(s) ROI_solve(p, solver = "alabama", start = s))
best_solution <- which.max(sapply(solus, solution, type = "objval"))  
round(solution(solus[[best_solution]]), 3)
```

## Example 14: Minimize 99\% CVaR wrt cardinality constraints {#cvarex2}

\begin{eqnarray*}
\min_{x\in \mathbb{R}^N, z\in \mathbb{R}^S, \gamma\in \mathbb{R}}& &\gamma + \frac{1}{(1-0.99)} \sum_{s = 1}^S p_s z_s\\
s.t.&&z_s \geq - r_s^\top x - \gamma,\\
&&z_s \geq 0,\\
&& \sum_{i=1}^N u_i\leq 6,\\ 
&& x_i - u_i \leq 0.
\end{eqnarray*}
where $u\in \mathbb{R}^N$are  binary auxiliary variables.

```{r}
tmp <- cvar_objective(djia2018, 0.99)
lp <- OP()

constraints(lp) <- rbind(
  tmp$constraint,
  budget_constraint(djia2018),
  cardinality_constraint(djia2018, dir = "<=", rhs = 6),
  use.names = TRUE)

obj <-c((tmp$objective)$L)
objective(lp) <- c(obj, double(NCOL(constraints(lp)) - length(obj)))

types(lp) <- rep("C",  NCOL(constraints(lp) ))
types(lp)[grep("z_card_aux", constraints(lp)$names)] <- "B"

(sol <- ROI_solve(lp, solver = "glpk"))

round(solution(sol)[1:30], 3)
```
