---
title: "How to use automatic differentiation with **ROI**?"
---

```{r}
Sys.setenv(ROI_LOAD_PLUGINS = FALSE)
library(ROI)
```

The following example is based on 
[`Automatic Differentiation in R by autodiffr` (Changcheng and Nash and Borchers 2018)](https://non-contradiction.github.io/autodiffr/articles/autodiffr_intro.html) which
is part of the **autodiffr** documentation.


First we load the  package.
```{r}
library(autodiffr)
ad_setup()
```

Users which want to use **autodiffr** with **ROI** have several options to do
so. The easiest and recommended way is just to make the gradient function 
with **autodiffr** and provide the gradient when constructing the optimization
problem.
```{r}
fun <- function(x) sum(x^2L)
grad <- makeGradFunc(fun)
hess <- makeHessianFunc(fun)

o <- OP(F_objective(F = fun, n = 3L, G = grad, H = hess))
s <- ROI_solve(o, solver = "nlminb", start = rnorm(3))
solution(s)
```

In general users can change the default differentiation function via `ROI_options`
but it is not recommended to change the default to **autodiffr**. Since not
every function works out of the box.

```{r}
fun0 <- function(x) {
    stopifnot(is.numeric(x), length(x) == 4L)
    det(matrix(x^c(1, 2, 1, 3), 2, 2))
}

x0 <- c(1.2, 1.4, 1.6, 1.8)
fun0(x0)
tryCatch(ad_grad(fun0, x0), error = function(e) e)
```

[Changcheng and Nash and Borchers (2018)](https://non-contradiction.github.io/autodiffr/articles/autodiffr_intro.html) find a way to resolve 
this problem by replacing `matrix` with `array` in the `fun0()` function.
```{r}
fun2 <- function(x) {
    stopifnot(is.numeric(x), length(x) == 4L)
    det(array(x^c(1, 2, 1, 3), c(2, 2)))
}

x0 <- c(1.2, 1.4, 1.6, 1.8)
fun2(x0)
ad_grad(fun2, x0)
```

However this example shows that a function (e.g. `fun0()`) which works perfectly 
fine in `R` and with **numDeriv**
```{r}
library(numDeriv)
numDeriv::grad(fun0, x0)
```
can cause errors in **autodiffr**. Therefore it is not recommended to set 
**autodiffr** as default option to derive the gradient (via `ROI_option`). 
Since this is likely to cause errors which are hard to debug.


